Nonlinear Dendritic Coincidence Detection for Supervised Learning

### 2021-04-06 notes/draft_article.tex
.

		In our work, we combined a phenomenological model predicting the output
		firing rate as a function of two streams of synaptic input (subsuming basal and apical input)
		with a BCM-like plasticity rule on basal synapses. LTP was induced in our model 
		if the postsynaptic activity surpassed a threshold level that was only achievable if 
		both basal and apical inputs were present. Therefore, we hypothesized that 
		this combination of neural activation and plasticity rules would lead to an
		increased correlation between basal and apical input.

		Furthermore, this temporal alignment could potentially facilitate apical inputs to act
		as top-down teaching signals, without the need for an explicit error-driven
		learning rule. Thus, we also tested our model in a simple linear 
		supervised classification task and compared it to the performance of simple
		point neuron combined with a BCM-plasticity rule on the basal inputs.

@@

		We implemented a BCM-like plasticity rule for the basal 
		synaptic weights given by the following update equation:
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{align}
			w_i(t+1) &= w_i(t) + \mu_w \left(x_{p,i}(t) - \tilde{x}_{p,i}(t)\right)
			y(t)\left[y(t) - \theta_y \right]
			\label{eq:bcm_plast} \\
			\tilde{x}_{p,i}(t+1) &= (1-\mu_{av})\tilde{x}_{p,i}(t) + \mu_{av}x_{p,i}(t+1)
		\end{align}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		While the classical BCM-rule dynamically modifies the 
		threshold $\theta_y$ to stabilize weights, we used a
		synaptic normalization constraint
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			w_i(t) \rightarrow \frac{w_i(t)}{\lVert \mathbf{w}(t)\rVert}
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		in each time step, where $\lVert \mathbf{w}(t)\rVert$ denotes the Euclidean
		norm of the synaptic weight vector. The threshold $\theta_y$ was
		chosen such that LTP is present if neural activity is in the
		high-activity regime (yellow area in Fig.~\ref{fig:comp_model}), while
		LTD is present in the low-activity regime 
		(given by $\alpha$, see Fig.~\ref{fig:comp_model}). Hence, we set
		$\theta_y = (1+\alpha)/2$.

		For comparative reasons, the point neuron model was equipped with the
		same plasticity rule for the proximal weights as \eqref{eq:bcm_plast}, 
		except that the threshold $\theta_y$ was dynamically adjusted as
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			\theta_y(t+1) = \left(1-\mu_{av}\right)\theta_y(t) + \mu_{av} y^2(t) \; , 
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		which is an implementation of a running average of the square activity,
		as used in the classic BCM-rule.

@@

		To be continued, see Fig.~\ref{fig:classification_accuracy_bcm}.

@@

@@

>>>>

		In our work, we combined a phenomenological model that generates the output
		firing rate as a function of two streams of synaptic input (subsuming basal and apical input)
		with Hebbian as well as BCM-like plasticity rules on basal synapses. We hypothesized that 
		this combination of neural activation and plasticity rules would lead to an
		increased correlation between basal and apical input.

		Furthermore, this temporal alignment could potentially facilitate apical inputs to act
		as top-down teaching signals, without the need for an explicit error-driven
		learning rule. Thus, we also tested our model in a simple linear 
		supervised classification task and compared it to the performance of a simple
		point neuron equipped with similar plasticity rules.

@@

		We implemented a Hebbian plasticity rule for the basal 
		synaptic weights given by the following update equation:
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{align}
			w_i(t+1) &= w_i(t) + \mu_w \left(x_{p,i}(t) - \tilde{x}_{p,i}(t)\right)
			\left(y(t) - \tilde{y} \right)
			\label{eq:bcm_plast} \\
			\tilde{x}_{p,i}(t+1) &= (1-\mu_{av})\tilde{x}_{p,i}(t) + \mu_{av}x_{p,i}(t+1) \\
			\tilde{y}(t+1) &= (1-\mu_{av})\tilde{y}(t) + \mu_{av}y(t+1)
		\end{align}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		Additionally, we used a
		synaptic normalization constraint
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			w_i(t) \rightarrow \frac{w_i(t)}{\lVert \mathbf{w}(t)\rVert}
			\label{eq:weight_norm}
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		in each time step, where $\lVert \mathbf{w}(t)\rVert$ denotes the Euclidean
		norm of the synaptic weight vector. 
%		The threshold $\theta_y$ was
%		chosen such that LTP is present if neural activity is in the
%		high-activity regime (yellow area in Fig.~\ref{fig:comp_model}), while
%		LTD is present in the low-activity regime 
%		(given by $\alpha$, see Fig.~\ref{fig:comp_model}). Hence, we set
%		$\theta_y = (1+\alpha)/2$.

		For comparative reasons, the point neuron model was equipped with the
		same plasticity rule for the proximal weights as \eqref{eq:bcm_plast}. 
%		except that the threshold $\theta_y$ was dynamically adjusted as
%		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%		\begin{equation}
%			\theta_y(t+1) = \left(1-\mu_{av}\right)\theta_y(t) + \mu_{av} y^2(t) \; , 
%		\end{equation}
%		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%		which is an implementation of a running average of the square activity,
%		as used in the classic BCM-rule.

@@

		Instead of Hebbian learning, we also considered a BCM-like
		learning rule for the basal weights \citep{Bienenstock1982,Intrator1992}.
		The form of the BCM-rule we consider here reads
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
		\Delta w_i \propto y\left(y - \theta_M\right) x_i - \epsilon w_i \; ,
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		where $\theta_M$ is a threshold defining a transition from LTP to LTD and
		$\epsilon$ is an optional decay term on the weights.
		In the variant introduced by \citet{Law1994}, the sliding threshold is simply
		the temporal average of the squared neural activity, 
		$\theta_M = \langle y^2 \rangle$. In practice, this would be calculated
		as a running average, thereby preventing the weights from growing 
		indefinitely.

		However, for our compartment model, we chose to explicitly set the
		threshold to be the mean value between the high- and low-activity regime
		in our compartment model, i.e.\ $\theta_M = (1+\alpha)/2$. By doing so, LTP was
		preferably induced if both basal and apical input was present at the same
		time. Furthermore, instead of the weight decay term, we chose to keep
		the weight normalization as introduced in \eqref{eq:weight_norm}.
		Obviously, for the point model, the reasoning behind our choice of
		$\theta_M$ did not apply. Still, to provide some level of comparability,
		we also ran simulations with a point model where the sliding threshold was
		calculated as a running average of $y^2$. Furthermore, we did not use
		weight normalization in this case, but chose to use a small weight decay term
		with $\epsilon = 0.1$. The results are shown in 
		Fig.~\ref{fig:classification_accuracy_bcm} (classification task) and 
		Fig.~\ref{fig:classification_accuracy_bcm} (Basal-Apical alignment). 
		While the accuracy of the classification for the point model
		was at most comparable to Hebbian learning, the BCM-like rule for the 
		compartment model significantly increased the accuracy for the tested
		parameter range (compare Fig.~\ref{fig:classification_accuracy}).

		Likewise, the compartment model also significantly benefited from
		the BCM rule in terms of basal-apical alignment, while only marginal
		improvements could be observed for the point model
		(compare Fig.~\ref{fig:corr_dimension_scaling_bcm} 
		with Fig.~\ref{fig:corr_dimension_scaling}).

@@

		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\includegraphics[width=\columnwidth]{corr_dimension_scaling_bcm}
			\caption{{\bf Alignment between Basal and Apical Input, BCM Rule.}
				Color encodes the Pearson correlation $\rho[I_p,I_d]$ for different
				number of orthogonal distraction directions $N_{dist}$ 
				and the corresponding scaling faction $s$ after training
				with a BCM-like rule. Compare to Fig.~\ref{fig:corr_dimension_scaling}.}
			\label{fig:corr_dimension_scaling_bcm}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

		To be continued
